Bulk Semantic Annotation with a Partially Known Knowledge Base\ednote{ Too \snify specific... }

### 2025-01-09 doc/cicm25-sysdesc/sysdesc.tex
more writing

\paragraph{Overview}
We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
describe and evaluate one concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
The paper ends with a brief conclusion (\cref{sec:conclusion}).

@@

\sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.
But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
    last paragraph doesn't fit here anymore, as it only motivates \snify
}

@@

\paragraph{Semantic Authoring in \sTeX}
How this can concretely work depends on the particular semantic framework, the specific
annotations we want to support, and the semantics embedded in the content commons. In the
\sTeX framework, where we semantically annotate formulae and technical terms with pointers
to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
in sentences and glyphs in formulae with the verbalizations and notations in the content
commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
formulae via notations has been presented by Vre{\v{c}}ar et al. in
\cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
annotation of technical terms via verbalizations.\ednote{move this in the next section}

\section{Existing and Envisioned Authoring Support for \sTeX/\ALeA}\label{sec:tools}

>>>>

\paragraph{Overview}
We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
describe and evaluate one concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
The paper ends with a brief conclusion (\cref{sec:conclusion}).
\ednote{rework this paragraph a bit}

@@

% \sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
% analyzes annotations on the fly, displays the underlying reference semantics, reports
% errors and redundancies, and even offers a concept search interface.
% But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
%     last paragraph doesn't fit here anymore, as it only motivates \snify
% }

@@

% \paragraph{Semantic Authoring in \sTeX}
% How this can concretely work depends on the particular semantic framework, the specific
% annotations we want to support, and the semantics embedded in the content commons. In the
% \sTeX framework, where we semantically annotate formulae and technical terms with pointers
% to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
% in sentences and glyphs in formulae with the verbalizations and notations in the content
% commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
% formulae via notations has been presented by Vre{\v{c}}ar et al. in
% \cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
% annotation of technical terms via verbalizations.\ednote{move this in the next section}

\section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
Semantic authoring can ``inherit'' tool support from both informal
and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.
But such features only support existing annotations and are, therefore,
of limited use when creating new annotations, which arguably is the
most time consuming aspects of semantic authoring.

How the annotation process can be supported depends on the particular semantic framework,
the specific annotations we want to support, and the semantics embedded in the content commons.
In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers to their definienda.
A tool for supporting semantic authoring of formulae via notations has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.
It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline|A\times B|) into abstract syntax trees based on
semantic macros defined in \sTeX, which can then replace the original formula
(e.g.\ \lstinline|\cart{A,B}| for the cartesian product of sets).
In the case of ambiguity, authors can pick the correct reading in a
graphical user interface that visualizes the different abstract syntax trees.

Efficient support for the annotation of technical terms in natural language
has been elusive so far.
\ednote{
    Move Hutter from conclusion here?
    Also: Plivelic}
